taxize - taxonomic search and retrieval in R

### 2013-07-05 inst/doc/taxize_withcode_plos.Rnw
added phylomatic code back + minor changes

The R language is the dominant language used by biologists (reference), and now has over 4,500 packages on the R package repository (CRAN) and more than 2,500 packages on other repositories to extend R. R is great for manipulating, visualizing and fitting statistical models to data. However, the key missing piece in R is the ability to get data from the internet within R. Getting data from the web will be increasingly common as more and more data gets moved to the cloud. Increasingly, data is available from the web via API's, or application programming interfaces. These are bits of code that allow computers to talk to one another using code that is not human readable, but is machine readable. Web APIs often define a number of methods that allow users to search for a species name, or retrieve the synonyms for a species name, for example. A further strength of APIs is that they are language agnostic, meaning that data can be consumed in almost any computing context, allowing users to interact with the web API without having to know the details of the code. Whereas, if data are stored in an Excel file, for example, the file can only be opened in a few programs.

@@

A number of data sources in taxize provide the capability to retrieve higher taxonomic names, but we will highlight two of the more useful ones: ITIS and NCBI. First, we'll search for two species, \emph{Abies procera} and \emph{Pinus contorta}.

@@

@@

\subsection*{What taxa are the children of my taxon of interest?}
If you aren't a taxonomic specialist on a particular taxon you likely don't know what children taxa are within a family, or within a genus. This task becomes especially unwieldy when there are a large number of taxa downstream. You can of course go to a website like Wikispecies (\cite{wikispecies}) or Encyclopedia of Life (\cite{eol}) to get downstream names. However, taxize provides an easy way to programatically search for downstream taxa, both for the Catalogue of Life (CoL; \cite{col}) and the Integrated Taxonomic Information System (\cite{itis}). Here is a short example using the CoL in which we want to find all the species within the genus \emph{Apis} (honey bees).

>>>>

The R language is the dominant language used by biologists (reference), and now has over 4,500 packages on the Comprehensive R Archive Network (CRAN) and more than 2,500 packages on other repositories to extend R. R is great for manipulating, visualizing and fitting statistical models to data. However, the key missing piece in R is the ability to get data from the internet within R. Getting data from the web will be increasingly common as more and more data gets moved to the cloud. Increasingly, data is available from the web via API's, or application programming interfaces. These are bits of code that allow computers to talk to one another using code that is not human readable, but is machine readable. Web APIs often define a number of methods that allow users to search for a species name, or retrieve the synonyms for a species name, for example. A further strength of APIs is that they are language agnostic, meaning that data can be consumed in almost any computing context, allowing users to interact with the web API without having to know the details of the code. Whereas, if data are stored in an Excel file, for example, the file can only be opened in a few programs.

@@

A number of data sources in taxize provide the capability to retrieve higher taxonomic names, but we will highlight two of the more useful ones: Integrated Taxonomic Information System (ITIS) \cite{itis} and National Center for Biotechnology Information (NCBI) \cite{federhen}. First, we'll search for two species, \emph{Abies procera} and \emph{Pinus contorta}.

@@

<<phylomaticphylogeny_disp, eval=FALSE>>=
taxa <- c("Poa annua", "Abies procera", "Helianthus annuus")
tree <- phylomatic_tree(taxa = taxa)
tree$tip.label <- capwords(tree$tip.label)
plot(tree, cex=1)
@

<<phylomaticphylogeny, eval=TRUE, dev='pdf', out.width='4in', out.height='4in', cache=TRUE, include=FALSE>>=
# Input the taxonomic names
taxa <- c("Poa annua", "Abies procera", "Helianthus annuus")

# Fetch the tree - the formatting of names and higher taxonmy is done within the function
tree <- phylomatic_tree(taxa = taxa)

# Captilize the species names
tree$tip.label <- capwords(tree$tip.label)

# Plot the tree
plot(tree, cex=1)
@

@@

\subsection*{What taxa are the children of my taxon of interest?}
If you aren't a taxonomic specialist on a particular taxon you likely don't know what children taxa are within a family, or within a genus. This task becomes especially unwieldy when there are a large number of taxa downstream. You can of course go to a website like Wikispecies \cite{wikispecies} or Encyclopedia of Life \cite{eol} to get downstream names. However, taxize provides an easy way to programatically search for downstream taxa, both for the Catalogue of Life (CoL) \cite{col} and the Integrated Taxonomic Information System \cite{itis}. Here is a short example using the CoL in which we want to find all the species within the genus \emph{Apis} (honey bees).
